---
title: "Data-clean"
output: html_document
---

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
```

## This document is for cleaning data from csv or Exel files.

### File 1

***File name: ***area_report_by_year.xlsx

***Citation: ***State Level Household Debt Statistics 2003-2024, Federal Reserve Bank of New York, April 2025	

***Source: ***New York Fed Consumer Credit Panel / Equifax	

***Note: ***The data here exclude US Territories like Guam and the US Virgin Islands, as well as certain other areas. In addition, they are subject to sampling variation. As such, national and state totals here may not match those reported in the Quarterly Report on Household Debt and Credit. 		
		

| Sheet                        | Title                                                                 | Notes                                   |
|-----------------------------|-----------------------------------------------------------------------|-----------------------------------------|
| `population`                | Number of Consumers in NY Fed Consumer Credit Panel                   | 5% of population (18+ with Equifax)     |
| `auto loan`                 | Auto Debt Balance per Capita                                          | 5% of the population                    |
| `credit card`               | Credit Card Debt Balance per Capita                                   | 5% of the population                    |
| `mortgage`                  | Mortgage Debt Balance per Capita (excluding HELOC)                    | 5% of the population                    |
| `student loan`              | Student Loan Debt Balance per Capita                                  | 1% of the population                    |
| `total`                     | Total Debt Balance per Capita                                         | 5% of the population                    |
| `auto loan delinquency`     | % of Auto Loan Balance 90+ Days Delinquent                            | 5% of the population                    |
| `credit card delinquency`   | % of Credit Card Balance 90+ Days Delinquent                          | 5% of the population                    |
| `mortgage delinquency`      | % of Mortgage Balance 90+ Days Delinquent                             | 5% of the population                    |
| `student loan delinquency`  | % of Student Loan Balance 90+ Days Delinquent (and in default)        | 1% of the population                    |


```{r, warning=FALSE, message=FALSE}
file_path <- "new-data/area_report_by_year.xlsx"

# There are many sheets so we should target it
# But auto has diffefent format so we will seperate it
target_sheets <- c(
"population",
"creditcard",
"mortgage",
"student loan",
"total",
"auto_delinq",
"creditcard_delinq",
"mortgage_delinq",
"studentloan_delinq"
)

auto <- read_excel(file_path, sheet = "auto", skip = 4, col_names = TRUE)

for (sheet_name in target_sheets) {
  
  df_name <- gsub(" ", "_", sheet_name)
  
  # The real data start from 9 row in the xlxs file. 
  df <- read_excel(file_path, sheet = sheet_name, skip = 8, col_names = TRUE)
  
  assign(df_name, df)
  #print(df_name)
}

# After load the file, it's nice to check but I will make it # to make the document short.
#population
#auto
#creditcard
#mortgage
#student_loan
#total
#auto_delinq
#creditcard_delinq
#mortgage_delinq
#studentloan_delinq
```
```{r, warning=FALSE, message=FALSE}
population_clean <- population %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    population_pre = pre,
    population_post = post
  )

head(population_clean)
```


```{r, warning=FALSE, message=FALSE}
creditcard_clean <- creditcard %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    creditcard_pre = pre,
    creditcard_post = post
  )

head(creditcard_clean)
```

```{r, warning=FALSE, message=FALSE}
auto_clean <- auto %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    auto_pre = pre,
    auto_post = post
  )

head(auto_clean)
```
```{r, warning=FALSE, message=FALSE}
mortgage_clean <- mortgage %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    mortgage_pre = pre,
    mortgage_post = post
  )

head(mortgage_clean)
```

```{r, warning=FALSE, message=FALSE}
studentloan_clean <- student_loan %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    studentloan_pre = pre,
    studentloan_post = post
  )

head(studentloan_clean)
```

```{r, warning=FALSE, message=FALSE}
total_clean <- total %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    total_pre = pre,
    total_post = post
  )

head(total_clean)
```

```{r, warning=FALSE, message=FALSE}
creditcard_delinq_clean <- creditcard_delinq %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    creditcard_delinq_pre = pre,
    creditcard_delinq_post = post
  )

head(creditcard_delinq_clean)
```

```{r, warning=FALSE, message=FALSE}
auto_delinq_clean <- auto_delinq %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    auto_delinq_pre = pre,
    auto_delinq_post = post
  )

head(auto_delinq_clean)
```

```{r, warning=FALSE, message=FALSE}
mortgage_delinq_clean <- mortgage_delinq %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    mortgage_delinq_pre = pre,
    mortgage_delinq_post = post
  )

head(mortgage_delinq_clean)
```

```{r, warning=FALSE, message=FALSE}
studentloan_delinq_clean <- studentloan_delinq %>%
  filter(!is.na(state)) %>%
  mutate(
    pre = rowMeans(select(., Q4_2017:Q4_2019), na.rm = TRUE),
    post = rowMeans(select(., Q4_2020:Q4_2022), na.rm = TRUE)
  ) %>%
  select(state, pre, post) %>%
  rename(
    studentloan_delinq_pre = pre,
    studentloan_delinq_post = post
  )

head(studentloan_delinq_clean)
```

```{r, warning=FALSE, message=FALSE}

#Join
NYFED_df <- population_clean %>%
  left_join(creditcard_clean, by = "state") %>%
  left_join(auto_clean, by = "state") %>%
  left_join(mortgage_clean, by = "state") %>%
  left_join(studentloan_clean, by = "state") %>%
  left_join(total_clean, by = "state") %>%
  left_join(creditcard_delinq_clean, by = "state") %>%
  left_join(auto_delinq_clean, by = "state") %>%
  left_join(mortgage_delinq_clean, by = "state") %>%
  left_join(studentloan_delinq_clean, by = "state")

head(NYFED_df)
```

```{r, warning=FALSE, message=FALSE}
# export to file but I will # because we already have sync folder with github
#write.csv(NYFED_df, "new-data/data-clean/NYFED.csv", row.names = FALSE)
```

### File 2
***Real GDP***

Annual GDP by State : https://apps.bea.gov/regional/downloadzip.htm
File name: SAGDP1__ALL_AREAS_1997_2024.csv


```{r, warning=FALSE, message=FALSE}
gdp_df <- read.csv("new-data/SAGDP1__ALL_AREAS_1997_2024.csv")
#gdp_df

gdp_real <- gdp_df %>%
  filter(Description == "Real GDP (millions of chained 2017 dollars) 1/")
head(gdp_real)

gdp_df2 <- gdp_real %>%
  select(GeoName, X2017, X2018, X2019, X2020, X2021, X2022) %>%
  rename(state = GeoName)

head(gdp_df2)
```
```{r, warning=FALSE, message=FALSE}
gdp_df3 <- gdp_df2 %>%
  select(state, X2017, X2018, X2019, X2020, X2021, X2022) %>%
  mutate(
    gdp_pre = rowMeans(select(., X2017, X2018, X2019), na.rm = TRUE),
    gdp_post = rowMeans(select(., X2020, X2021, X2022), na.rm = TRUE)
  )

gdp_df4 <- gdp_df3 %>% select(state, gdp_pre, gdp_post)

gdp_df4 <- gdp_df4 %>%
  filter(state != "United States")

head(gdp_df4)
```

```{r, warning=FALSE, message=FALSE}
# We need mapping with code 
state_codes <- read.csv("https://raw.githubusercontent.com/jasonong/List-of-US-States/master/states.csv")

#state_codes

gdp_df5 <- gdp_df4 %>%
  left_join(state_codes, by = c("state" = "State"))

gdp_final <- gdp_df5 %>% 
  select(Abbreviation, gdp_pre, gdp_post) %>% 
  rename(State = Abbreviation )

head(gdp_final)
```
```{r, warning=FALSE, message=FALSE}
# export to file but I will # because we already have sync folder with github
# write.csv(gdp_final, "new-data/data-clean/GDP.csv", row.names = FALSE)
```

### File 3
***Personal Consumption Expenditures (PCE)***

Personal Consumption Expenditures (PCE) by State: https://apps.bea.gov/regional/downloadzip.htm

File name: SAPCE1__ALL_AREAS_1997_2023.csv

```{r, warning=FALSE, message=FALSE}
pce <- read.csv("new-data/SAPCE1__ALL_AREAS_1997_2023.csv")
#pce
pce_df <- pce %>%
  filter(Description == "Personal consumption expenditures ")

head(pce_df)

pce_df2 <- pce_df %>%
  select(GeoName, X2017, X2018, X2019, X2020, X2021, X2022) %>%
  rename(state = GeoName)

#head(pce_df2)

pce_df2 <- pce_df2 %>%
  filter(state != "United States")

head(pce_df2)
```

```{r, warning=FALSE, message=FALSE}
pce_df3 <- pce_df2 %>%
  select(state, X2017, X2018, X2019, X2020, X2021, X2022) %>%
  mutate(
    pce_pre = rowMeans(select(., X2017, X2018, X2019), na.rm = TRUE),
    pce_post = rowMeans(select(., X2020, X2021, X2022), na.rm = TRUE)
  )

pce_df4 <- pce_df3 %>% select(state, pce_pre, pce_post)
pce_df4

pce_df5 <- pce_df4 %>%
  left_join(state_codes, by = c("state" = "State"))

pce_final <- pce_df5 %>% 
  select(Abbreviation, pce_pre, pce_post) %>% 
  rename(State = Abbreviation )

head(pce_final)
```

```{r, warning=FALSE, message=FALSE}
# export to file but I will # because we already have sync folder with github
#write.csv(pce_final, "new-data/data-clean/PCE.csv", row.names = FALSE)
```

### File4
**Annual Personal Income**

Annual Personal Income by State: https://apps.bea.gov/regional/downloadzip.htm
File name: SAINC1__ALL_AREAS_1929_2024.csv

```{r, warning=FALSE, message=FALSE}
income <- read.csv("new-data/SAINC1__ALL_AREAS_1929_2024.csv")

income_df1<- income %>%
  filter(str_trim(Description) == "Per capita personal income (dollars) 2/")

#income_df1

income_df2 <- income_df1 %>%
  select(GeoName, X2017, X2018, X2019, X2020, X2021, X2022) %>%
  rename(state = GeoName) %>%
  mutate(
    income_pre = rowMeans(select(., X2017, X2018, X2019), na.rm = TRUE),
    income_post = rowMeans(select(., X2020, X2021, X2022), na.rm = TRUE)
  ) %>%
  filter(state != "United States")

income_df2
```

```{r, warning=FALSE, message=FALSE}
income_df3 <- income_df2 %>%
  left_join(state_codes, by = c("state" = "State"))

income_final <- income_df3%>% 
  select(Abbreviation, income_pre, income_post) %>% 
  rename(State = Abbreviation)

head(income_final)
```
```{r, warning=FALSE, message=FALSE}
# export to file but I will # because we already have sync folder with github
#write.csv(income_final, "new-data/data-clean/INCOME.csv", row.names = FALSE)
```

### File5

Unemployment rate by state
```{r, warning=FALSE, message=FALSE}

unemploy <- read_excel("new-data/unemployment rate.xlsx", range = "A1:DR53")
names(unemploy)[-1] <- as.character(as.Date(as.numeric(names(unemploy)[-1]), origin = "1899-12-30"))

pre_df <- unemploy %>%
  select(State, matches("2017|2018|2019"))
#pre_df 

post_df <- unemploy %>%
  select(State, matches("2020|2021|2022"))

# there were some chr
pre_df_numeric <- pre_df %>%
  mutate(across(-State, ~ as.numeric(.)))
# there were some chr
post_df_clean <- post_df %>%
  mutate(across(-State, as.numeric))  

# Average
pre_avg <- pre_df_numeric %>%
  mutate(unemp_pre = rowMeans(select(., -State), na.rm = TRUE)) %>%
  select(State, unemp_pre)

post_avg <- post_df_clean %>%
  mutate(unemp_post = rowMeans(select(., -State), na.rm = TRUE)) %>%
  select(State, unemp_post)

unemploy2 <- pre_avg %>%
  inner_join(post_avg, by = "State")

unemploy3 <- unemploy2%>%
  left_join(state_codes, by = c("State" = "State"))

unemploy_final <- unemploy3 %>% 
  select(Abbreviation, unemp_pre, unemp_post) %>% 
  rename(State = Abbreviation)


unemploy_final
```

```{r, warning=FALSE, message=FALSE}
# export to file but I will # because we already have sync folder with github
#write.csv(unemploy_final , "new-data/data-clean/UNEMPLOY.csv", row.names = FALSE)
```

### File6

Housing data: https://www.zillow.com/research/data/
File name: State_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month.csv

```{r, warning=FALSE, message=FALSE}
house <- read_csv("new-data/State_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month.csv")

house1 <- house %>%
  rename(State = RegionName) %>%
  select(-StateName, -RegionID, -SizeRank, -RegionType)
```

```{r, warning=FALSE, message=FALSE}
house_long <- house1 %>%
  pivot_longer(cols = -State, names_to = "Date", values_to = "ZHVI")
#house_long

house_long1 <- house_long %>%
  filter(str_sub(Date, 1, 4) %in% c("2017", "2018", "2019", "2020", "2021", "2022"))

#house_long1

house_long2 <- house_long1 %>%
  mutate(year = as.numeric(str_sub(Date, 1, 4)))

house_avg <- house_long1 %>%
  group_by(State, year) %>%
  summarise(mean_zhvi = mean(ZHVI, na.rm = TRUE), .groups = "drop")

#house_avg

house2 <- house_avg %>%
  pivot_wider(names_from = year, values_from = mean_zhvi)

house2
```
```{r, warning=FALSE, message=FALSE}
house3 <- house2 %>%
  mutate(
    zhvi_pre_avg = rowMeans(select(., `2017`, `2018`, `2019`), na.rm = TRUE),
    zhvi_post_avg = rowMeans(select(., `2020`, `2021`, `2022`), na.rm = TRUE),
    zhvi_pre_change = ((`2019` - `2017`) / `2017`)*100,
    zhvi_post_change = ((`2022` - `2020`) / `2020`)*100
  ) %>%
  select(State, zhvi_pre_avg, zhvi_post_avg, zhvi_pre_change, zhvi_post_change)

#house3

house4 <- house3 %>%
  left_join(state_codes, by = c("State" = "State"))

house_final <- house4  %>%
  select(Abbreviation, zhvi_pre_avg, zhvi_post_avg, zhvi_pre_change, zhvi_post_change) %>%
  rename(State = Abbreviation)

house_final 
```
```{r}
# export to file but I will # because we already have sync folder with github
# write.csv(house_final, "new-data/data-clean/HOUSING.csv", row.names = FALSE)
```

