---
title: "Statistic"
author: "Nabiha Chowdhury & Anna Hyunjung Kim"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)

library(tidyverse)
library(DT)
library(patchwork)
library(ggthemes)
library(plotly)
library(gapminder)
library(grid)
library(ggplot2)
library(gridExtra)

library(DescTools)
library(car)
library(rpart)
library(rpart.plot)
```

```{r, warning=FALSE, message=FALSE}

df_stat <- read_csv("data/data-clean/FINAL.csv")
df_stat
```
## ANOVA {.tabset}


| Variable           | ANOVA p-value | Post-Hoc Interpretation                                       | Shapiro p-value | Levene p-value | Equal Variance |
|--------------------|----------------|----------------------------------------------------------------|------------------|------------------
| **gdp_post**        | 2.40e-08       | All three groups significantly different from each other     | 0.798            | 0.273            
| **income_post**     | 0.039          | Low group < Medium & High groups (no difference between Medium & High) | 0.299            | 0.344            
| **unemp_post**      | 0.122          | No statistically significant differences between groups      | 0.679            | 0.605            
| **total_post**      | 0.00016        | Differences exist; Post-hoc details required (likely Low vs High) | 0.401            | 0.138            
| **creditcard_post** | 0.333          | No statistically significant differences between groups      | 0.873            | 0.188            

### Population
```{r}
df_stat$pop_group <- cut(df_stat$population_post,
                    breaks = quantile(df_stat$population_post, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                    labels = c("Small", "Medium", "Large"),
                    include.lowest = TRUE)

pop_anova <- aov(zhvi_post_change ~ pop_group, data = df_stat)
summary(pop_anova)

PostHocTest(pop_anova, method = "lsd")

ggplot(df_stat, aes(x = pop_group, y = zhvi_post_change, fill = pop_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Post-COVID Housing Price Change by Population Group",
       x = "Population Group",
       y = "zhvi_post_change")

# residual analysis to check the assumptions
shapiro.test(pop_anova$residuals)  

# Levene's test for homogeneity of variance
leveneTest(zhvi_post_change ~ factor(pop_group), data = df_stat)
```

### Income
```{r}
df_stat$income_group <- cut(df_stat$income_post,
                       breaks = quantile(df_stat$income_post, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                       labels = c("Low", "Medium", "High"),
                       include.lowest = TRUE)

income_anova <- aov(zhvi_post_change ~ income_group, data = df_stat)
summary(income_anova)

PostHocTest(income_anova, method = "lsd")

ggplot(df_stat, aes(x = income_group, y = zhvi_post_change, fill = income_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Housing Price Change by Income Group",
       x = "Income Group",
       y = "zhvi_post_change")

# residual analysis to check the assumptions
shapiro.test(income_anova$residuals)  

# Levene's test for homogeneity of variance
leveneTest(zhvi_post_change ~ factor(income_group), data = df_stat)
```
### GDP

```{r}
df_stat$gdp_group <- cut(df_stat$gdp_post,
                         breaks = quantile(df_stat$gdp_post, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                         labels = c("Low", "Medium", "High"),
                         include.lowest = TRUE)

gdp_anova <- aov(zhvi_post_change ~ gdp_group, data = df_stat)
summary(gdp_anova)

PostHocTest(gdp_anova, method = "lsd")

shapiro.test(gdp_anova$residuals)

leveneTest(zhvi_post_change ~ factor(gdp_group), data = df_stat)

ggplot(df_stat, aes(x = gdp_group, y = zhvi_post_change, fill = gdp_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Housing Price Change by GDP Group",
       x = "GDP Group",
       y = "zhvi_post_change")
```

### Unemployment

```{r}
df_stat$unemp_group <- cut(df_stat$unemp_post,
                           breaks = quantile(df_stat$unemp_post, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                           labels = c("Low", "Medium", "High"),
                           include.lowest = TRUE)

unemp_anova <- aov(zhvi_post_change ~ unemp_group, data = df_stat)
summary(unemp_anova)

PostHocTest(unemp_anova, method = "lsd")
shapiro.test(unemp_anova$residuals)
leveneTest(zhvi_post_change ~ factor(unemp_group), data = df_stat)

ggplot(df_stat, aes(x = unemp_group, y = zhvi_post_change, fill = unemp_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Housing Price Change by Unemployment Group",
       x = "Unemployment Group",
       y = "zhvi_post_change")
```

### Total Dept
```{r}
df_stat$total_group <- cut(df_stat$total_post,
                           breaks = quantile(df_stat$total_post, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                           labels = c("Low", "Medium", "High"),
                           include.lowest = TRUE)

total_anova <- aov(zhvi_post_change ~ total_group, data = df_stat)
summary(total_anova)

PostHocTest(total_anova, method = "lsd")
shapiro.test(total_anova$residuals)
leveneTest(zhvi_post_change ~ factor(total_group), data = df_stat)

ggplot(df_stat, aes(x = total_group, y = zhvi_post_change, fill = total_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Housing Price Change by Total Debt Group",
       x = "Total Debt Group",
       y = "zhvi_post_change")
```

### Credit Card
```{r}
df_stat$creditcard_group <- cut(df_stat$creditcard_post,
                                breaks = quantile(df_stat$creditcard_post, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
                                labels = c("Low", "Medium", "High"),
                                include.lowest = TRUE)

creditcard_anova <- aov(zhvi_post_change ~ creditcard_group, data = df_stat)
summary(creditcard_anova)

PostHocTest(creditcard_anova, method = "lsd")
shapiro.test(creditcard_anova$residuals)
leveneTest(zhvi_post_change ~ factor(creditcard_group), data = df_stat)

ggplot(df_stat, aes(x = creditcard_group, y = zhvi_post_change, fill = creditcard_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Housing Price Change by Credit Card Debt Group",
       x = "Credit Card Debt Group",
       y = "zhvi_post_change")
```


## Decision Tree
```{r}
tree_model <- rpart(zhvi_post_change ~ zhvi_post_avg + population_post + gdp_post + income_post + unemp_post + total_post + creditcard_post
                    ,data = df_stat
                    ,method = "anova"
                    ,control = rpart.control(cp = 0.0001,minsplit = 10, minbucket = 4)) # it's not class because our house price is 

rpart.plot(tree_model,
           extra = 101,      
           fallen.leaves = TRUE,
           main = "Decision Tree")

tree_model$variable.importance

node_assignments <- tree_model$where

tree_nodes <- data.frame(State = df_stat$State, Node = node_assignments)

#tree_nodes
datatable(tree_nodes, 
          filter = "top",
          caption = "state")
```

## Clustering

### 6 vars
```{r}
clusters <- c("population_post", "gdp_post", "income_post", "unemp_post", "total_post", "zhvi_post_avg")
cluster_z <- scale(df_stat[, clusters] )

set.seed(612)
kmeans <- kmeans(cluster_z, centers = 5)

#kmeans

df_stat$cluster <- as.factor(kmeans$cluster)

ggplot(df_stat, aes(x = cluster, y = zhvi_post_change, fill = cluster)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Cluster for Post COVID",
       x = "Cluster Group",
       y = "zhvi_post_change")

cluster_nodes <- data.frame(State = df_stat$State, Cluster = df_stat$cluster)

#cluster_nodes
split(cluster_nodes$State, cluster_nodes$Cluster)
```

```{r}
cluster_summary <- aggregate(df_stat[, clusters], by = list(Cluster = df_stat$cluster), FUN = mean)
cluster_summary[ , -1] <- round(cluster_summary[ , -1], 2)
print(cluster_summary)
```

### 4 vars

```{r}
clusters2 <- c("population_post", "income_post", "total_post", "zhvi_post_avg")
cluster_z2 <- scale(df_stat[, clusters2])

set.seed(612)
kmeans2 <- kmeans(cluster_z2, centers = 6)
df_stat$cluster2 <- as.factor(kmeans2$cluster)

ggplot(df_stat, aes(x = cluster2, y = zhvi_post_change, fill = cluster2)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Clustering: GDP + Income + Population",
       x = "Cluster",
       y = "Housing Price Change (%)")
```

```{r}
cluster_nodes2 <- data.frame(State = df_stat$State, Cluster = df_stat$cluster2)
split(cluster_nodes2$State, cluster_nodes2$Cluster)
```

```{r}
cluster_summary2 <- aggregate(df_stat[, clusters2], by = list(Cluster = df_stat$cluster2), FUN = mean)
cluster_summary2[ , -1] <- round(cluster_summary2[ , -1], 2)
print(cluster_summary2)
```


